### 2013-11-11 src2/thesis.tex
update

CPUs and GPUs get more parallel resources with each new generation. The question of how to 
program these processors efficiently arises. We are after efficiency both in programmer 
productivity and in application performance. Using embedded languages allows us to experiment
with what abstractions to present to the programmer at relatively little effort.

Obsidian is an embedded language for general purpose programming of GPUs. We try to strike 
a balance between high level, productivity increasing abstractions and low-level 
control needed for performance. The Obsidian programming model mirrors the GPU architecture 
and the programmer is constrained into writing GPU-friendly code. By supplying hierarchy level 
polymorphic library functions, these constraints does not feel obtrusive. Obsidian programs are 
compiled into CUDA C code. This compilation is based on a simple and elegant monad reification 
technique.

@@

The work behind paper H was done by me as a hobby project and my aspirations were not 
high. I just wanted to experiment with some embedded languages techniques that we had 
been avoiding (had no need for) in Obsidian. Mary discovered what I was doing and applied 
the right amount of pressure, leading to us encompassing much more of ArBB functionality than 
I had in mind. The paper's text is joint work; most examples as well as benchmarking are my work,
but the sparse matrix vector multiplication example is entirely Mary's.

@@

Computers today are increasingly parallel. Each new generation of general purpose processors
(CPUs) and of graphics processors (GPUs), brings more parallelism. The question 
of how to program these computers most efficiently, both from a productivity and 
a performance point of view, arises. Explicit parallel programming is hard and automatic 
means of parallelisation, while giving a sequential programming model to the programmer, 
often misses the target when it comes to performance. The book ``Structured Parallel 
Programming'' by Michael McCool et al argues for the need of explicit parallel programming 
models~\citet{STRUCTURED}.

The embedded language approach to programming language implementation is beneficial, 
since we are not entirely sure of how to program current and future highly parallel 
computers. Implementing an embedded language allows us more rapid prototyping of ideas. 
We can experiment with what abstractions to expose to the programmer, without much 
effort. Other benefits of embedded languages often mentioned are: making use 
of the host language's library functions, if they are general enough. The host language 
can be used as a powerful macro language for the embedded language. An embedded 
language also provides a more leveled learning curve to programmers already familiar 
with the host language. For these programmers, using an embedded language is more like  
learning about a new library.

This thesis focuses on data-parallelism, as opposed to task or control parallelism.
Data-parallelism is applicable where the input data can be divided into chunks
and operated upon independently. Data-parallelism is scalable; the potential 
for parallelism increases with the problem size. Two different embedded languages 
for data-parallel programming are presented in this thesis: Obsidian, for 
general purpose computations on GPUs and EmbArBB for parallel programming across 
platforms, CPU, GPU and other accelerators.

Graphics Processors (GPUs) trade programmer convenience for increased parallelism. 
For example, where CPUs dedicate large portions of chip area to caches, branch 
prediction and instruction level parallelism (ILP), GPUs put additional cores. 
This comes at a price of increased programmer effort. GPUs have programmer managed 
shared memories, conditionals are troublesome from a performance 
perspective and memory access patterns can make or break the performance of an 
application. With Obsidian, we try to find the right level of abstraction for 
programming GPUs. We try to strike a balance between low-level control of details 
that influence performance and high-level abstractions to increase productivity and 
make programs more terse. The end goal with Obsidian is a language for writing 
high performance GPU programs that compare well to code written by CUDA~\citet{CUDA}  
(a C-dialect for GPU programming) experts, provided that both programmers have 
similar GPU expertise. The Obsidian programmer would however do so in a more 
composable manner. Obsidian programs are compiled into CUDA.

Obsidian is geared towards the programmer who is interested in GPU hardware. Making 
good use of Obsidian requires that the programmer is willing to make an effort 
to learn about GPUs and how to program them efficiently. However, another case 
is a programmer who does not want to invest that effort but wants to write a high-level 
parallel program and have it run reasonably well on different parallel architectures. 
In this case, a language with even higher level of abstraction is preferable. It then 
becomes the job of a more advanced compiler to turn that into code for the targeted 
platforms. The EmbArBB language targets this programmer. EmbArBB exposes a set of high-level 
operations on arrays, such as {\tt map}, {\tt reduce} and {\tt scan}. The programmer 
has no power over the parallel strategy that should be used to implement these 
operations but instead relies on a JIT-compiler (Just In Time-compiler). EmbArBB is a 
Haskell embedding of Intel ArBB that was released as an embedded language in C++. 
EmbArBB relies entirely on the ArBB system for code generation and optimisation.

@@

In 2006 NVIDIA released their first Compute Unified Device Architecture (CUDA) 
GPU. With CUDA, NVIDIA also made it a lot easier for programmers to use GPUs
for general purpose programming. A dialect of C, called CUDA C, was released 
specifically to simplify the implementation of general purpose computations on GPUs.

@@

Before going into programming of GPUs, it is important to have some background. The 
GPU programming model exposed by CUDA very much mirrors the underlying hardware. Some 
of the details that make GPU programming hard is more apparent when looking at the 
underlying hardware.

@@

Each MP of the GPU can manage a large number of threads; on today's GPUs up to 
2048 threads can run on a single MP. The GPU schedules threads in groups of 32, called 
{\em Warps}, that are executed in lock-step (SIMD style execution). Threads are 
also divided into {\em Blocks}; the threads within a block can communicate using the 
shared memory. The maximum block size is currently 1024 threads. Since 
the number of threads in a block can be higher than the warp size, a barrier 
synchronisation mechanism exists to ensure that threads within a block 
can communicate safely via the shared memory. Blocks are also grouped into a grid that 
is the collection of blocks executing the same program.

@@

\subsection{GPU programming} 
\label{GPUPROGRAMMING}

The CUDA programming model reflects the hierarchical architecture of the GPU. The 
programmer writes what are called \emph{kernels}; in CUDA jargon a kernel refers to a sequential 
C program that is parameterised on a thread's identity. The kernel program is executed 
in parallel across all the threads of a block and many such blocks of threads can execute 
in parallel on the available MPs. Kernels may use shared memory that is local to a block; 
the programmer needs to decompose the algorithm and data in such a way that communication 
will be local to the blocks.

@@

In Obsidian, the same kernel (the local computation) is implemented as follows:

@@

\noindent The {\tt vecAdd} program takes two {\tt SPull} input arrays and gives one 
as result. The Obsidian {\tt vecAdd} is more general than the CUDA version. From it 
we can generate CUDA kernels for any element type that supports the {\tt (+)} operation. 
In Haskell we know this is true for types that are member of the {\tt Num} typeclass.

There are some important differences between CUDA and Obsidian programs. One is that with Obsidian 
we generate kernels for fixed size local computations. This is to avoid doing out-of-bounds check 
which would mean conditionals in the code. Conditionals are problematic from a performance 
point of view. Another difference is that in Obsidian the programmer specifies the entire 
computation, including how it is mapped over the blocks of the GPU. This is done implicitly 
in CUDA, based on the indexing arithmetic used in the kernel and the number of blocks and 
threads provided at kernel launch. This means that the Obsidian programmer needs to 
write a program that performs this mapping:

@@

There are some superficial differences between the code generated using Obsidian and the 
handwritten example above. One such difference is that the index computation is not shared. 
Obsidian relies on the CUDA compiler to discover such sharing via a Common Subexpression 
Elimination (CSE) optimisation pass. Another difference is that Obsidian has inserted a 
conditional that only executes the body for certain blocks. Discussion of the causes for 
that conditional is postponed until a later section~\ref{sec:OBSARRAYS}. A real 
difference is that the kernel is generated for a fixed size, here there is a constant (32) 
where in the handwritten code we used {\tt blockDim.x}. If those differences are set 
aside, the two kernels are identical. Figure~\ref{fig:OBSIDIANHOST1} shows how the 
{\tt vecAddG} kernel can be launched from within Haskell, using Obsidian's rudimentary 
support for launching GPU computations.

@@

In Obsidian, local computations can also use shared memory. The Obsidian arrays 
we have seen in the examples, {\tt Pull} and {\tt Push} arrays do not directly 
correspond to actual data in memory. Pull and push arrays, rather represent ways 
to compute arrays. More information about these array representations is available 
in section~\ref{sec:OBSARRAYS}. The Obsidian programmer can store intermediate 
arrays in shared memory by using a {\tt force} operation. 
%The 
%{\tt force} operation computes an array and stores it into shared memory.

@@

\noindent This code uses a shared memory array called {\tt sbase}. There 
is also a call to {\tt \_\_syncthreads} to ensure that all the writes have 
completed before proceeding. In this case the synchronisation is not really
necessary since all threads belong to the same warp. There is a {\tt unsafeforce} 
operation that naively removes synchronisations if the array being forced is short 
enough. More advanced analysis could also be performed to remove synchronisations 
whenever communication is entirely warp local. We do not apply that optimisation, 
but work on warp level computations is in progress and would put this ability 
in the hands of the programmer.

@@

\item {\bf Minimise data transfers between host and device:} The bandwidth between device memory 
and GPU is much higher than the bandwidth between host memory and the GPU device via the PCIe bus.
This means that, first of all one must evaluate if the computation to be performed is significant 
enough to warrant a transfer to the GPU. If the computation is not arithmetic intensive, it may be 
better to do the work on the CPU. However, if the data is already in device memory and 
the operation that is about to perform on it is more efficiently handled by the CPU, it may still 
be better to let the GPU do the work. In other words, compute near the data unless transferring it 
pays off.

@@

\item {\bf Ensure that global memory accesses are coalesced:} Certain access patterns into 
GPU device memory can be {\em coalesced} (combined into few memory transactions). Unfortunately 
what these access patterns are varies somewhat depending on the GPU. For a typical model of 
GPU, the rule is that concurrent memory accesses (by threads within a warp) will be coalesced 
to as many transactions as cache lines touched (128b l1 cache lines). This means that strided 
accesses from within a warp are not favoured.

Inputs to an Obsidian program are pull arrays. Pull arrays can be arbitrarily permuted; 
the permutation used will decide the access pattern. If this method can be applied or 
not partly also depends on the algorithm being implemented.

\item {\bf Minimise use of global memory:} Accessing global device memory can take as 
many as 400 to 600 clock cycles if the data is not cached. The programmer should 
use shared memory as much as possible to avoid redundant loading from global memory.
Shared memory can also be used to coalesce reads from global memory. Shared memory is 
a limited resource in the MP, so if a kernel uses very much shared memory it means 
fewer such blocks can be active on the MP.

@@

\item {\bf Use a multiple of 32 threads per block:} Groups of 32 threads, warps, are the 
scheduled unit on a GPU. If a warp accesses memory (and will be waiting for it arrive during 
600 cycles) it will be swapped out and another warp will take its place. This means that 
the number of available warps decides how well memory latency can be hidden. A block that is 
not a multiple of 32 leads to wasting GPU resources; some processing elements will stand 
idle during the execution of the not full warps.

@@

Sometimes divergent code cannot be avoided. However, it is important not to have threads 
diverge unnecessarily. When using pull arrays alone this was often a problem; with the 
addition of push arrays the situation improved. Using push arrays some divergent computations 
can be explicitly split up into phases, each phase with all threads following the 
same execution path. This means that fewer threads are required to do the same work.

@@

In Obsidian it is impossible for a barrier synchronisation to appear within diverging 
conditional code. Conditionals can be introduced either by an \\{\tt ifThenElse} function 
at the element level or by a {\tt Cond} statement that is an operation in the {\tt Program} 
monad; we have seen an example of the program monad in section~\ref{sec:GPUPROGRAMMING}. 
The {\tt Program} monad is parameterised on the GPU hierarchy level. There are 
thread programs ({\tt TProgram}), block programs ({\tt BProgram}) and grid programs 
({\tt GProgram}). the {\tt Cond} conditional choose between two thread programs while a 
synchronisation can only appear in an Obsidian block program. More information about programs 
and the GPU hierarchy is available in section~\ref{sec:OBSIMP}.

@@

The list above provides some very general guidelines on what to think about when 
writing GPU kernels. The lesson is that fast memory near the processing element 
should be preferred over slow memory far away. However, this comes at the effort 
of decomposing computation and data in a way that makes this possible.

@@

Obsidian is implemented as an embedded language using Haskell as host. Obsidian 
is a compiled embedded language; a very good paper that is highly relevant to our 
approach, is ``Compiling Embedded Languages'' by Conal Elliott et al~\citet{COMPILEEDSL}. 
This paper describes how to embed a compiler backend and generate efficient code 
in some target language. One key aspect of this approach is that the embedded 
language is a library of functions that create and compose Abstract Syntax Trees 
(ASTs). Obsidian (chapter \ref{chap:GPUProgramming}) as well as EmbArBB 
(chapter \ref{chap:ArBB}) are implemented in a similar way. An embedding that builds 
ASTs is called a deep embedding. Shallow embeddings, on the other hand, do not build 
ASTs. In ``Combining Deep and Shallow Embedding for EDSL'' Josef Svenningsson and 
Emil Axelsson combine the two methods, shallow and deep, in order to simplify the 
AST data type (fewer constructors) and make more extensible embedded languages 
\citet{DEEPSHALLOW}. Obsidian also uses a combination of shallow 
and deep embeddings.

The embedded language approach has become popular to use as a way of raising the 
level of abstraction in fields where the default is to use low-level languages. 
For example, Lava and Wired~\citet{LAVA,Wired} raises the level of abstraction in the 
fields of hardware design and verification and Feldspar does the same in the area of  
digital signal processing~\citet{FELDSPAR2010}. For GPU programming, there are numerous 
embedded approaches using various host languages. Two languages for GPU programming 
embedded in Haskell are  Accelerate and Nikola \citet{ACCELERATEDAMP11, NIKOLA}. 
And there is Intel ArBB, embedded in C++, with the goal of being parallel across 
platforms \citet{ARBB2011}. Related work and information about the tools, libraries 
and language that make the wider context surrounding Obsidian can be found in 
section~\ref{sec:relatedwork}.

@@

Obsidian uses a deeply embedded {\tt Program} data type that represents GPU Kernels. 
A CUDA GPU has a hierarchy of parallel resources. At the bottom there are threads, each 
executing sequential programs. There are groups of threads (of 32) called {\em Warps} that 
execute in lock-step. Warps are the scheduled unit of work on a GPU. There are {\em Blocks} 
of threads, a set of threads that run as a group and share local (shared) memory. And 
lastly there is a {\em Grid} of blocks that specifies the total number of threads involved 
in a computation (Number\_of\_Blocks * Number\_of\_Threads). The deeply embedded program 
data type of Obsidian models this hierarchy by parameterising programs on a hierarchy 
level type parameter (the  {\tt t} parameter below).

@@

The {\tt Thread} type is level {\tt Zero} ({\tt type Thread = Zero}). Then {\tt Block} is 
{\tt Step Thread} and {\tt Grid} is {\tt Step Block}. Currently {\tt Warp} is not included 
in the hierarchy leading to warp programming being a special case. One benefit a warp has is 
that threads can communicate via shared memory without using synchronisation primitives. There 
is no programmer support for warp level programming in CUDA (as there is for block level 
programs). The programmer needs to take care to ensure that the communication patterns are 
within a warp and is then free to remove synchronisations. If warps were to be put in the 
program level hierarchy of Obsidian, its place would be between the thread and block level. 
This would force the programmer to go via warps when putting together a block computation even 
if the communication pattern is not suited for that, leading to inconvenience with no gain.

@@

Scalars and operations on scalars are represented by an expression data type ({\tt Exp a}) in 
Obsidian. This is another example of a deep embedding; the expression data type contains 
constructors for literal values, arithmetic operations, and conditionals.

@@

The approach taken to embed the scalar language is very similar to what is described 
in ``Compiling Embedded Languages''~\citet{COMPILEEDSL}; but we use a Generalised Algebraic 
Datatype (GADT) to obtain typed expressions. An alternative to the GADT, is to use phantom 
types. Earlier versions of Obsidian (section~\ref{sec:paperB}) used phantom types.

@@

Pull arrays are parameterised on both element ({\tt a}) and length ({\tt s}) type. The length 
parameter is used to be able to represent arrays with both static (known at Haskell runtime) 
or dynamic (length represented by an expression, an unknown length). When creating block 
level computations, we want to know the exact lengths in order to avoid generating code 
riddled with conditionals concerning array lengths. It is possible to write kernels that 
are not specialised for a certain size. This is accomplished by adding bounds checks. Since 
conditionals  are problematic from a performance point of view, fixed size kernels are recommended 
when performance is crucial. Dynamic lengths are used for grid level computations, this allows 
a local computation of a fixed size to be run in parallel over arrays a multiple of the size that 
the local computation handles. In generated code this is currently visible as a conditional 
on {\tt blockIdx.x}. However, this is still considered work in progress; the conditional is 
only really necessary for Obsidian kernels that have more than one output array and the outputs
are of different length. 
%Conditionals are problematic on GPUs and 
%while general programs that run for any size is possible when performance is crucial a fixed 
%size is recommended.

@@

Another benefit of pull arrays are that they are parallel. Simply put, early Obsidian was pull 
arrays in addition to a way to compute the array and store its elements to memory 
(called {\tt sync} in early versions of Obsidian and later {\tt force}). A very 
direct way to compile a pull array to a GPU is to launch as many threads as there are 
elements in the array and apply the indexing function to the thread id. Now each thread 
will have computed an element of the array and can store that to memory.

@@

Push arrays were added to Obsidian as a complement to pull arrays. Pull arrays are 
good because they are so simple to understand and easy to parallelise. However, certain
operations on pull arrays yield code that is not suitable for GPU execution. Concatenation 
or interleaving of pull arrays result in conditionals in the indexing function. The 
interleaving case, which is worst, leads to diverging branching in every warp. If branches 
diverge, the GPU will turn off the processing elements taking one path and 
allow the others to progress; then turning to the other branch. The computation of the 
branches is serialised and compute resources are wasted.

@@

When working with pull arrays, the consumer of such an array decides how to 
iterate over the array and what elements to compute. This is done applying the 
push array function to those indices that are interesting to the consumer. Push 
arrays work in the opposite way. The Push array function (the result program) contain 
the iteration schema; the consumer of the push only decides what write function 
to supply. The iteration schema used is decided upon when creating a push arrays. 
The code below is an example of how to create a push array from a pull array.

>>>>

CPUs and GPUs get more parallel resources with each new generation. The question of how to 
efficiently program these processors arises. We are after efficiency both in programmer 
productivity and in application performance. Using embedded languages allows us to experiment
with what abstractions to present to the programmer at relatively little effort.

Obsidian is an embedded language for general purpose programming of GPUs. We try to strike 
a balance between high level, productivity increasing abstractions and low-level 
control needed for performance. The Obsidian programming model mirrors the GPU architecture 
and the programmer is constrained into writing GPU-friendly code. Hierarchy level 
polymorphic library functions are supplied to make these constraints feel less obtrusive. 
Obsidian programs are compiled into CUDA C code. This compilation is based on a simple 
and elegant monad reification technique.

@@

The work behind paper H was done by me as a hobby project and my aspirations were not 
high. I just wanted to experiment with some embedded language techniques that we had 
been avoiding (had no need for) in Obsidian. Mary discovered what I was doing and applied 
the right amount of pressure, leading to us encompassing much more of ArBB functionality than 
I had in mind. The paper's text is joint work; most examples as well as benchmarking are my work,
but the sparse matrix vector multiplication example is entirely Mary's.

@@

Computers today are increasingly parallel. Each new generation of general purpose processors
(CPUs) and of graphics processors (GPUs) brings more parallelism. The question 
of how to program these computers most efficiently, both from a productivity and 
a performance point of view, arises. Explicit parallel programming is hard and automatic 
means of parallelisation, while giving a sequential programming model to the programmer, 
often misses the target when it comes to performance. The book ``Structured Parallel 
Programming'' by Michael McCool et al argues the need for explicit parallel programming 
models~\citet{STRUCTURED}.

The embedded language approach to programming language implementation is beneficial, 
since we are not entirely sure of how to program current and future highly parallel 
computers. Implementing an embedded language allows us more rapid prototyping of ideas. 
We can experiment with what abstractions to expose to the programmer, without much 
effort. There are many other benefits of embedded languages: making use 
of the host language's library functions (if they are general enough), the host language 
can be used as a powerful macro language for the embedded language, providing a more 
leveled learning curve to programmers already familiar with the host language. 
For programmers who are already proficient in the host language, learning an embedded 
language is more like picking up a new library.

This thesis focuses on data-parallelism, as opposed to task or control parallelism.
Data-parallelism is applicable where the input data can be divided into chunks
and operated upon independently. Data-parallelism is scalable; the potential 
for parallelism increases with the problem size. Two different embedded languages 
for data-parallel programming are presented in this thesis: Obsidian, for 
general purpose computations on GPUs and EmbArBB for parallel programming across 
platforms, such as CPUs, GPUs and other accelerators.

Graphics Processors trade programmer convenience for increased parallelism. 
For example, where CPUs dedicate large portions of chip area to caches, branch 
prediction and instruction level parallelism (ILP), GPUs put additional cores. 
This comes at a price of increased programmer effort. GPUs have programmer managed 
shared memories, conditionals are troublesome from a performance 
perspective and memory access patterns can make or break the performance of an 
application. With Obsidian, we try to find the right level of abstraction for 
programming GPUs. We try to strike a balance between low-level control of details 
that influence performance and high-level abstractions to increase productivity and 
make programs more terse. The end goal with Obsidian is a language for writing 
high performance GPU programs that compare well to code written by CUDA~\citet{CUDA}  
(a C-dialect for GPU programming) experts, provided that both programmers have 
similar GPU expertise. The Obsidian programmer would however write programs in a more 
composable manner. Obsidian programs are compiled into CUDA.

Obsidian is geared towards the programmer who is interested in GPU hardware. Making 
good use of Obsidian requires that the programmer is willing to make an effort 
to learn about GPUs and how to program them efficiently. However, another case 
is a programmer who does not want to invest that effort but wants to write a high-level 
parallel program and have it run reasonably well on different parallel architectures. 
In this case, a language with even higher level of abstraction is preferable. It then 
becomes the job of a more advanced compiler to turn that into code for the targeted 
platforms. The EmbArBB language targets this programmer. EmbArBB exposes a set of high-level 
operations on arrays, such as {\tt map}, {\tt reduce} and {\tt scan}. The programmer 
has no power over the parallel strategy that should be used to implement these 
operations but instead relies on a JIT-compiler (Just In Time-compiler). EmbArBB is a 
Haskell embedding of Intel ArBB~\citet{ARBB2011} that was released as an embedded language 
in C++. EmbArBB relies entirely on the ArBB system for code generation and optimisation.

@@

In 2006, NVIDIA released their first Compute Unified Device Architecture (CUDA) 
GPU. With CUDA, NVIDIA also made it a lot easier for programmers to use GPUs
for general purpose programming~\citet{wwwcuda}. A dialect of C, called CUDA C, was released 
specifically to simplify the implementation of general purpose computations on GPUs.

@@

Before going into programming of GPUs, it is important to have some background. The 
GPU programming model exposed by CUDA very much mirrors the underlying hardware. Some 
of the details that make GPU programming hard are more apparent when looking at the 
underlying hardware.

@@

Each MP of the GPU can manage a large number of threads; on today's GPUs up to 
2048 threads can run on a single MP. The GPU schedules threads in groups of 32, called 
{\em Warps}, that are executed in lock-step (SIMD style execution). Threads are 
also divided into {\em Blocks}; the threads within a block can communicate using the 
shared memory. The maximum block size is currently 1024 threads. Threads within 
a warp can communicate via the shared memory without using any synchronisation 
primitive. However, if communication takes place across warps synchronisation is 
necessary. A barrier synchronisation mechanism exists to ensure that all threads within a block 
have reached the same position in the code. Blocks are also grouped into a grid that 
is the collection of blocks executing the same program.

@@

\subsection{GPU programming} 
\label{sec:GPUPROGRAMMING}

The CUDA programming model reflects the hierarchical architecture of the GPU. The 
programmer writes what are called \emph{kernels}; in CUDA jargon, a kernel refers to a sequential 
C program that is parameterised on a thread's identity. The kernel program is executed 
in parallel across all the threads of a block and many such blocks of threads can execute 
in parallel on the available MPs. Kernels may use shared memory that is local to a block; 
the programmer needs to decompose the algorithm and data in such a way that communication 
will be local to the blocks.

@@

\noindent In Obsidian, the same kernel (the local computation) is implemented as follows:

@@

\noindent The {\tt vecAdd} program takes two {\tt SPull} input arrays and gives one 
as result. The Obsidian {\tt vecAdd} is more general than the CUDA version. From it 
we can generate CUDA kernels for any element type that supports the {\tt (+)} operation. 
In Haskell we know that types that belong to the {\tt Num} typeclass has a {\tt (+)} operation.

There are some important differences between CUDA and Obsidian programs. One is that with Obsidian 
we generate kernels for fixed size local computations. This is to avoid doing out-of-bounds checks 
which would mean conditionals in the code. Conditionals are problematic from a performance 
point of view. Another difference is that in Obsidian the programmer specifies the entire 
computation, including how it is mapped over the blocks of the GPU. This is done implicitly 
in CUDA, based on the indexing arithmetic used in the kernel and the number of blocks and 
threads provided at kernel launch. This means that the Obsidian programmer needs to 
write a program that performs this mapping:

@@

There are some superficial differences between the code generated using Obsidian and the 
handwritten example above. One such difference is that the index computation is not shared. 
Obsidian relies on the CUDA compiler to discover such sharing via a Common Subexpression 
Elimination (CSE) optimisation pass. Another difference is that Obsidian has inserted a 
conditional that only executes the body for certain blocks. Discussion of the causes for 
that conditional is postponed until section~\ref{sec:OBSARRAYS}. A real 
difference is that the kernel is generated for a fixed size; here there is a constant (32) 
where in the handwritten code we used {\tt blockDim.x}. If those differences are set 
aside, the two kernels are identical. Figure~\ref{fig:OBSIDIANHOST1} shows how the 
{\tt vecAddG} kernel can be launched from within Haskell, using Obsidian's rudimentary 
support for launching GPU computations.

@@

In Obsidian, local computations can also use shared memory. The Obsidian arrays 
we have seen in the examples, {\tt Pull} and {\tt Push} arrays do not directly 
correspond to actual data in memory. Pull and push arrays rather represent ways 
to compute arrays. More information about these array representations is available 
in section~\ref{sec:OBSARRAYS}. The Obsidian programmer can store intermediate 
arrays in shared memory by using a {\tt force} operation. 
%The 
%{\tt force} operation computes an array and stores it into shared memory.

@@

\noindent This code uses a shared memory array called {\tt sbase}. There 
is also a call to \\ {\tt \_\_syncthreads} to ensure that all the writes have 
completed before proceeding. In this case the synchronisation is not really
necessary since all threads belong to the same warp. There is an {\tt unsafeforce} 
operation that naively removes synchronisations if the array being forced is short 
enough. More advanced analysis could also be performed to remove synchronisations 
whenever communication is entirely warp local. We do not apply that optimisation, 
but work on warp level computations is in progress and would put this ability 
in the hands of the programmer.

@@

\item {\bf Minimise data transfers between host and device:} The bandwidth between device memory 
and GPU is much higher than the bandwidth between host memory and the GPU device via the PCIe bus.
This means that, first of all, one must evaluate if the computation to be performed is significant 
enough to warrant a transfer to the GPU. If the computation is not arithmetic intensive, it may be 
better to do the work on the CPU. However, if the data is already in device memory and 
the operation that is about to be performed on it is more efficiently handled by the CPU, it 
may still be better to let the GPU do the work. In other words, compute near the data unless 
transferring it pays off.

@@

\item {\bf Ensure that global memory accesses are coalesced:} Certain access patterns into 
GPU device memory can be {\em coalesced} (combined into few memory transactions). Unfortunately 
what these access patterns are varies somewhat depending on the GPU. For a typical model of 
GPU, the rule is that concurrent memory accesses (by threads within a warp) will be coalesced 
to as many transactions as cache lines touched (128 Byte l1 cache lines). This means that strided 
accesses from within a warp are not favoured.

Inputs to an Obsidian program are pull arrays. Pull arrays can be arbitrarily permuted; 
the permutation used will decide the access pattern. If global memory accesses can 
be coalesced or not partly also depends on the algorithm being implemented.

\item {\bf Minimise use of global memory:} Accessing global device memory can take as 
many as 400 to 600 clock cycles if the data is not cached. The programmer should 
use shared memory as much as possible to avoid redundant loading from global memory.
Shared memory can also be used to coalesce reads from global memory. Shared memory is 
a limited resource in the MP, so if a kernel uses very much shared memory, 
fewer such blocks can be active on the MP.

@@

\item {\bf Use a multiple of 32 threads per block:} Groups of 32 threads, warps, are the 
scheduled unit on a GPU. If a warp accesses memory (and will be waiting for it arrive during 
600 cycles) it will be swapped out and another warp will take its place. This means that 
the number of available warps decides how well memory latency can be hidden. A block that is 
not a multiple of 32 leads to wasting GPU resources; some processing elements will stand 
idle during the execution of the incomplete warps.

@@

Sometimes divergent code cannot be avoided. However, it is important not to have threads 
diverge unnecessarily. When using pull arrays alone, this was often a problem; with the 
addition of push arrays, the situation improved. Using push arrays, some divergent computations 
can be explicitly split up into phases, each phase with all threads following the 
same execution path. This means that fewer threads are required to do the same work.

@@

In Obsidian, it is impossible for a barrier synchronisation to appear within diverging 
conditional code. Conditionals can be introduced either by an \\{\tt ifThenElse} function 
at the element level or by a {\tt Cond} statement that is an operation in the {\tt Program} 
monad; we have seen an example of the program monad in section~\ref{sec:GPUPROGRAMMING}. 
The {\tt Program} monad is parameterised on the GPU hierarchy level. There are 
thread programs ({\tt TProgram}), block programs ({\tt BProgram}) and grid programs 
({\tt GProgram}). the {\tt Cond} conditional chooses between two thread programs while a 
synchronisation can only appear in an Obsidian block program. More information about programs 
and the GPU hierarchy is available in section~\ref{sec:OBSIMP}.

@@

The list above provides some very general guidelines on what to think about when 
writing GPU kernels. The lesson is that fast memory near the processing element 
should be preferred over slow memory far away. Favouring fast memory demands the effort 
of decomposing computation and data in a way that makes this possible.

@@

Obsidian is implemented as an embedded language using Haskell as host. Obsidian 
is a compiled embedded language; a very good paper that is highly relevant to our 
approach is ``Compiling Embedded Languages'' by Conal Elliott et al~\citet{COMPILEEDSL}. 
This paper describes how to embed a compiler backend and generate efficient code 
in some target language. One key aspect of this approach is that the embedded 
language is a library of functions that create and compose Abstract Syntax Trees 
(ASTs). Obsidian (chapter \ref{chap:GPUProgramming}) as well as EmbArBB 
(chapter \ref{chap:ArBB}) are implemented in a similar way. An embedding that builds 
ASTs is called a deep embedding. Shallow embeddings, on the other hand, do not build 
ASTs. In ``Combining Deep and Shallow Embedding for EDSL'' Josef Svenningsson and 
Emil Axelsson combine the two methods, shallow and deep, in order to simplify the 
AST data type (fewer constructors) and make more extensible embedded languages 
\citet{DEEPSHALLOW}. Obsidian also uses a combination of shallow 
and deep embeddings.

The embedded language approach has become popular to use as a way of raising the 
level of abstraction in fields where the default is to use low-level languages. 
For example, Lava and Wired~\citet{LAVA,Wired} raise the level of abstraction in the 
fields of hardware design and verification and Feldspar does the same in the area of  
digital signal processing~\citet{FELDSPAR2010}. For GPU programming, there are numerous 
embedded approaches using various host languages. Two languages for GPU programming 
embedded in Haskell are  Accelerate and Nikola \citet{ACCELERATEDAMP11, NIKOLA}. 
And there is Intel ArBB, embedded in C++, with the goal of being parallel across 
platforms \citet{ARBB2011}. Related work and information about the tools, libraries 
and languages that make the wider context surrounding Obsidian can be found in 
section~\ref{sec:relatedwork}.

@@

Obsidian uses a deeply embedded {\tt Program} data type that represents GPU Kernels. 
A CUDA GPU has a hierarchy of parallel resources. At the bottom there are threads,  
executing sequential programs. There are groups of threads (of 32) called {\em Warps} that 
execute in lock-step. Warps are the scheduled unit of work on a GPU. There are {\em Blocks} 
of threads, a set of threads that run as a group and share local (shared) memory. And 
lastly there is a {\em Grid} of blocks that specifies the total number of threads involved 
in a computation (Number\_of\_Blocks * Number\_of\_Threads). The deeply embedded program 
data type of Obsidian models this hierarchy by parameterising programs on a hierarchy 
level type parameter (the  {\tt t} parameter below).

@@

The {\tt Thread} type is level {\tt Zero} ({\tt type Thread = Zero}). Then {\tt Block} is 
{\tt Step Thread} and {\tt Grid} is {\tt Step Block}. Currently {\tt Warp} is not included 
in the hierarchy, leading to warp programming being a special case. One benefit a warp has is 
that threads can communicate via shared memory without using synchronisation primitives. There 
is no programmer support for warp level programming in CUDA (as there is for block level 
programs). The programmer needs to take care to ensure that the communication patterns are 
within a warp and is then free to remove synchronisations. If warps were to be put in the 
program level hierarchy of Obsidian, its place would be between the thread and block level. 
This would force the programmer to go via warps when putting together a block computation even 
if the communication pattern is not suited for that, leading to inconvenience with no gain.

@@

Scalars and operations on scalars are represented by an expression data type \\ ({\tt Exp a}) in 
Obsidian. This is another example of a deep embedding; the expression data type contains 
constructors for literal values, arithmetic operations, and conditionals.

@@

The approach taken to embed the scalar language is very similar to what is described 
in ``Compiling Embedded Languages''~\citet{COMPILEEDSL}; but we use a Generalised Algebraic 
Datatype (GADT) to obtain typed expressions. An alternative to using a GADT is to use phantom 
types, which earlier versions of Obsidian did (section~\ref{sec:paperB}).

@@

Pull arrays are parameterised on both element ({\tt a}) and length ({\tt s}) type. The length 
parameter is used to be able to represent arrays with static (known at Haskell runtime) 
or dynamic (length represented by an expression, an unknown length). When creating block 
level computations, we want to know the exact lengths in order to avoid generating code 
riddled with conditionals concerning array lengths. In CUDA, it is possible to write kernels that 
are not specialised for a certain size. This is accomplished by adding bounds checks. Since 
conditionals  are problematic from a performance point of view, fixed size kernels are recommended 
when performance is crucial. In Obsidian, we have chosen to disallow implementation of kernels 
that operate on different sizes locally.  Dynamic lengths are used for grid level computations; 
this allows a local computation of a fixed size to be run in parallel over arrays a multiple
of the size that the local computation handles. In generated code, this is currently visible 
as a conditional on {\tt blockIdx.x}. However, this is still considered work in progress; the 
conditional is only really necessary for Obsidian kernels that have more than one output array 
and the outputs are of different length. 
%Conditionals are problematic on GPUs and 
%while general programs that run for any size is possible when performance is crucial a fixed 
%size is recommended.

@@

Another benefit of pull arrays is that they are parallel. Simply put, early Obsidian was pull 
arrays in addition to a way to compute the array and store its elements to memory 
(called {\tt sync} in early versions of Obsidian and later {\tt force}). A very 
direct way to compile a pull array to a GPU is to launch as many threads as there are 
elements in the array and apply the indexing function to the thread id. Now each thread 
will have computed an element of the array and can store that to memory.

@@

Push arrays were added to Obsidian as a complement to pull arrays. Pull arrays are 
good because they are so simple to understand and easy to parallelise. However, certain
operations on pull arrays yield code that is not suitable for GPU execution. Concatenation 
and interleaving of pull arrays results in conditionals in the indexing function. The 
interleaving case, which is worst, leads to diverging branching in every warp. If branches 
diverge, the GPU turns the processing elements taking one path off and 
allow the others to progress; then turn to the other branch. The computation of the 
branches is serialised and compute resources are wasted.

@@

When working with pull arrays, the consumer of such an array decides how to 
iterate over the array and what elements to compute. This is done by applying the 
push array function to those indices that are interesting to the consumer. Push 
arrays work in the opposite way. The Push array (the result program) contains 
the iteration schema; the consumer of the push only decides what write function 
to supply. The iteration schema used is decided upon when creating a push array. 
The code below is an example of how to create a push array from a pull array.
